Tidal to tell: \\ Does the Milky Way affect Sculptor and Ursa Minor?

### 2025-03-14 thesis/introduction.tex
markdown filters

\subsection{Background \& Past Work}\label{background-past-work}

@@

@@

\begin{figure}
\centering
\includegraphics{figures/scl_selection.png}
\caption{Sculptor selection criteria.}\label{fig:sculptor_selection}
\end{figure}

We use J+24 data, updated PM from the MV 2020a catalogue.

J+24 select members using a multi-component Baysian algorithm.

\begin{itemize}
\tightlist
\item
  First, stars with poor Gaia astrometry, color excess nearby
  parallaxes, and proper motions \textgreater{} 10 mas/yr are removed
\item
  Stars are assigned a likelihood based on the location on the CMD
  (assuming)
\end{itemize}

Simple cuts, such as selecting PM, CMD based on purely geometric cuts
reveal a similar density profile.

One caveat is the original algorithm takes spatial position into
account. When deriving a density profile, this assumption may influence
the derived density profile, especially when the galaxy density is
fainter than the background of similar appearing stars. To remidy this
and estimate where the background begins to take over, we also explore a
cut based on the likelihood ratio of only the CMD and PM components.

\subsection{Searches for tidal tails}\label{searches-for-tidal-tails}

\begin{figure}
\centering
\includegraphics{/Users/daniel/Library/Application Support/typora-user-images/image-20250313132102118.png}
\caption{image-20250313132102118}
\end{figure}

@@

Given a set of probable members, the density profile is given by
EQUATION, where we use \ldots{}

@@

  \begin{itemize}
  \tightlist
  \item
    The Sérsic index, \(n\), is a measure of the deviation from an
    exponential. Exponentials have \(n=1\), whereas more extended dwarf
    galaxies will have higher \(n\)
  \end{itemize}
\item
  To better estimate the uncertanties due to unknown galaxy properties
  and flexibility in the likelihood cut, we can
\end{itemize}

\begin{figure}
\centering
\includegraphics{figures/density_profiles_medley.png}
\caption{Density profiles}\label{fig:sculptor_observed_profiles}
\end{figure}

Dark Matter is one of the biggest open questions in modern astronomy.
Hello fig.~\ref{fig:sculptor}.

\subsection{Comparison of the Classical
dwarfs}\label{comparison-of-the-classical-dwarfs}

@@

@@

>>>>

\section{Background \& Past Work}\label{background-past-work}

@@

\subsubsection{Sculptor}\label{sculptor}

\begin{itemize}
\tightlist
\item
  sestito+23a
\item
  tolstoy+22
\item
  battaglia+09
\item
  arroyo-polonio+22
\end{itemize}

\begin{table*}[t]
\centering
\caption{Measured properties of Sculptor}
\label{tbl:Measured-properties-of-Sculptor}
\begin{tabular}{lll}
\toprule
parameter & value & Source\\
\midrule
$\alpha$ & $15.0183 \pm 0.0012$˚ & M+18\\
$\delta$ & $-33.7186 \pm 0.00072$˚ & M+18\\
distance & $83.2 \pm 2$ kpc & Tran+22\\
$\mu_\alpha \cos \delta$ & $0.099 \pm 0.002 \pm 0.017$ mas yr$^{-1}$ & MV20a\\
$\mu_\delta$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$ mas yr$^{-1}$ & MV20a\\
RV & $111.03 \pm 0.23$ & This work\\
$\sigma_v$ & $9.61\pm0.16$ & This work\\
$r_h$ & $12.33 \pm 0.05$ arcmin & MV20*\\
ell & $0.36 \pm 0.01$ & M+18\\
PA & $92\pm1$ & M+18\\
$M_V$ & $-10.82\pm0.14$ & M+18\\
$\Upsilon_\star$ & $1.5 \pm 0.3$ & assumed\\
\bottomrule
\end{tabular}
\end{table*}

@@

\begin{figure}
\centering
\includegraphics{figures/scl_selection.png}
\caption{Sculptor selection criteria. This will be a super long caption
to test the figure list.}\label{fig:sculptor_selection}
\end{figure}

We use J+24 data. J+24 select members using a multi-component Baysian
algorithm:

\begin{itemize}
\item
  Remove stars with poor astrometry or photometry, no colour excess
  (\citet{lindegren+2018} equation C.2), 3\(\sigma\) consistency of
  measured parallax with dwarf distance (near zero with
  \citet{lindegren+2018} zero-point correction), and absolute RA and Dec
  proper motions less than 10\(\,{\rm mas\ yr^{-1}}\).
\item
  Spatial likelihood based on a double exponential
  \(\Sigma_\star \propto e^{-r/r_s} + B\,e^{-r/r_{\rm outer}}\) where
  the inner scale radius is fixed.
\item
  Stars are assigned a likelihood based on the location on the CMD
  (using Padova isochrones including an intrinsic 0.1 CMD width in
  colour convolved with colour and distance modulus)
\item
  Background KDE density maps for the CMD and PM are constructed using
  the other quality-selected stars outside of \(5R_h\), where the
  satellite density would be orders of magnitude less than the
  background (even in the presence of extended tidal features).
\item
  Likelihoods normalized to unity to represnt a PDF

  Membership probabilities are then given by
\end{itemize}

\[
P_{\rm sat} = \frac{f_{\rm sat}{\cal L}_{\rm sat}}{f_{\rm sat}{\cal L}_{\rm sat} + (1-f_{\rm sat}){\cal L}_{\rm bg}} = \frac{1}{1 + \frac{(1-f_{\rm sat}){\cal L}_{\rm bg}}{f_{\rm sat}{\cal L}_{\rm sat} }}
\]

where \(f_{\rm sat}\) is the fraction of stars belonging to the system
inside the given field, \({\cal L}_{\rm sat}\) is the likelihood of a
star belonging to the satellite, and \({\cal L}_{\rm bg}\) is the
likelihood of the star belonging to the background. Each likelihood is
calculated as a product of the CMD, PM, and spatial likelihoods: \[
{\cal L} = {\cal L}_{\rm space}\ {\cal L}_{\rm PM}\ {\cal L}_{\rm CMD}
\]

Not shown here, we explore simple cuts of the stars, using absolute cuts
in parallax, proper motions, and the CMD. The results are similar to the
nospace model.

The above formula suggests that a cut in \(P_{\rm sat}\) is equivalent
to the cut in likelihoods \[
\frac{{\cal L}_{\rm sat}}{{\cal L}_{\rm bg}} > \frac{(1-f_{\rm sat})/f_{\rm sat}}{1/P_{\rm sat}- 1}
\]

Note that if we remove the spatial component of the likelihood, then
\(f_{\rm sat}\) represents a global normalization

\subsubsection{Searches for tidal tails}\label{searches-for-tidal-tails}

\begin{figure}
\centering
\includegraphics{figures/scl_tidal_tails.png}
\caption{Density profiles}\label{fig:sculptor_tidal_tails}
\end{figure}

\begin{itemize}
\item
  The above figure shows the distribution of member stars (orange), PM
  \& CMD only selected stars (blue) and all stars (passing quality cuts,
  black)
\item
  There are no apparent overdensities in the PM \& CMD only selected
  stars to suggest the presence of a tidal tail
\item
  This means that at least at the level of where the background density
  dominates, we can exclude models which produce tidal tails brighter
  than a density of
  \(\Sigma_\star \approx 10^{-2}\,\text{Gaia-stars\ arcmin}^{-2} \approx 10^{-6} \, {\rm M_\odot\ kpc^{-2}}\)
  (TODO assuming a distance of \ldots{} and stellar mass of \ldots).
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  How well do we know the density profiles?
\item
  What uncertainties affect derived density profiles?
\item
  Can we determine if Gaia, structural, or algorithmic systematics
  introduce important errors in derived density profiles?
\end{itemize}

J+24's algorithm takes spatial position into account, assuming either a
one or two component exponential density profile. When deriving a
density profile, this assumption may influence the derived density
profile, especially when the galaxy density is fainter than the
background of similar appearing stars. To remedy this and estimate where
the background begins to take over, we also explore a cut based on the
likelihood ratio of only the CMD and PM components. This is in essence
assuming that the spatial position of a star contains no information on
it's membership probability (a uniform distribution like the background)

To incorporate the structural uncertainties and robustly model the
sampling uncertainty, we construct the following bootstrap model

\begin{itemize}
\item
  Centre is varied by a centring error, estimated from the standard
  normal error of the positions plus the systematic shift of the mean
\item
  Position angle and ellipticity are sampled from a normal distribution
  given the reported uncertainties
\item
  \(f_{\rm sat}\) is sampled from \ldots{}

  \textbf{TODO}: Look into normalization of Likelihoods and check how
  \(f_{\rm sat}\) matters. If \(f_{\rm sat}\) is related to the
  normalizations of fg / bg densities, and other likelihoods are
  area-normalized to 1, then this makes life much easier.

  Test if psat weighted density profiles are similar

  Save MC density profile outputs
\end{itemize}

\begin{figure}
\centering
\includegraphics{figures/scl_density_methods.png}
\caption[Density profiles]{add simple cuts \& delve to convince
more\ldots{}}\label{fig:sculptor_observed_profiles}
\end{figure}

\section{Comparison of the Classical
dwarfs}\label{comparison-of-the-classical-dwarfs}

@@

  \begin{itemize}
  \tightlist
  \item
    The Sérsic index, \(n\), is a measure of the deviation from an
    exponential. Exponentials have \(n=1\), whereas more extended dwarf
    galaxies will have higher \(n\)
  \end{itemize}
\item
  To better estimate the uncertainties due to unknown galaxy properties
  and flexibility in the likelihood cut, we can
\end{itemize}

@@

\subsection{Radial Velocity
Measurements}\label{radial-velocity-measurements}

@@

\textbf{TODO}: Use updated density profiles (nospace) with uncertainties
included to MCMC fit Sérsic profiles to every dwarf galaxy.
